---
title: "Cancer Incidence by Washington Health Service Area"
format: html
editor: visual
---

```{r}
library(pak)
pak("getwilds/cancerprof@fix-incidence")
pak("duckdb")
pak("duckplyr")
pak("DBI")
pak("dbplyr")
pak("sf")
pak("ggplot2")
pak("dplyr")
pak("RColorBrewer")
```

## Collect Incidence Data (by cancer type)

```{r}
library(duckdb)
library(duckplyr)
library(DBI)
library(dbplyr)
library(sf)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

```{r}
# Create a new DuckDB database
con <- dbConnect(duckdb::duckdb(), "cancer-incidence-wa-hsa.duckdb")
```

```{r}
get_incidence_db_name <- function(cancer, race, sex, age, stage, year){
  ## all cancer sites = all
  ## bladder = bladder
  ## brain & ons = brain
  ## colon & rectum = colon
  ## esophagus = esophagus
  ## kidney & renal pelvis = kidney
  ## leukemia = leukemia
  ## liver & bile duct = liver
  ## lung & bronchus = lung
  ## melanoma of the skin = melanoma
  ## non-hodgkin lymphoma = lymphoma
  ## oral cavity & pharynx = oral
  ## pancreas = pancreas
  ## stomach = stomach
  ## thyroid = thyroid
  c_name = "allsites" # all cancer sites
  if (cancer == "bladder" ||
     cancer == "esophagus" ||
     cancer == "pancreas" ||
     cancer == "stomach" ||
     cancer == "thyroid" ||
     cancer == "leukemia") {
    c_name = cancer
  } else if (cancer == "brain & ons") {
    c_name = "brain"
  } else if (cancer == "colon & rectum") {
    c_name = "colon"
  } else if (cancer == "kidney & renal pelvis") {
    c_name = "kidney"
  } else if (cancer == "liver & bile duct") {
    c_name = "liver"
  } else if (cancer == "lung & bronchus") {
    c_name = "lung"
  } else if (cancer == "melanoma of the skin") {
    c_name = "melanoma"
  } else if (cancer == "non-hodgkin lymphoma") {
    c_name = "lymphoma"
  } else if (cancer == "oral cavity & pharynx") {
    c_name = "oral"
  } 

  ## All Races (includes Hispanic) = all
  ## White (non-Hispanic) = white
  ## Black (non-Hispanic) = black
  ## American Indian / Alaska Native (non-Hispanic) = native
  ## Asian / Pacific Islander (non-Hispanic) = asian
  ## Hispanic (Any Race) = hisp
  r_name = "allraces" # All Races (includes Hispanic)
  if (race == "White (non-Hispanic)") {
    r_name = "white"
  } else if (race == "Black (non-Hispanic)") {
    r_name = "black"
  } else if (race == "American Indian / Alaska Native (non-Hispanic)") {
    r_name = "native"
  } else if (race == "Asian / Pacific Islander (non-Hispanic)") {
    r_name = "asian"
  } else if (race == "Hispanic (Any Race)") {
    r_name = "hisp"
  }
  
  ## both sexes = b
  ## male = m
  ## female = f
  sx_name = "bothsexes" # both sexes
  if (sex == "male") {
    sx_name = "male"
  } else if (sex == "female") {
    sx_name = "female"
  } 

  ## all ages = all
  ## ages <50 = <50
  ## ages 50+ = 50+
  ## ages <65 = <65
  ## ages 65+ = 65+
  ## ages <15 = 15
  ## ages <20 = 20
  a_name = "allages" # all ages
  if (age == "ages <50") {
    a_name = "<50"
  } else if (age == "ages 50+") {
    a_name = "50+"
  } else if (age == "ages <65") {
    a_name = "<65"
  } else if (age == "ages 65+") {
    a_name = "65+"
  } else if (age == "ages <15") {
    a_name = "15"
  } else if (age == "ages <20") {
    a_name = "20"
  }
  
  ## all stages = all
  ## late stage (regional & distant) = late
  st_name = "allstages" # all stages
  if (stage == "late stage (regional & distant)") {
    st_name = "latestage"
  }
  
  ## latest 5 year average = avg
  ## latest single year (us by state) = one
  y_name = "5yr" # latest 5 year average
  if (stage == "latest single year (us by state)") {
    y_name = "1yr"
  }

  out = paste0("wa_hsa_", c_name, "_", r_name, "_", sx_name, "_", a_name, "_", st_name, "_", y_name)
  return(out)
}
```

```{r}
get_incidence_df <- function(cancer, race, sex, age, stage, year){
  out = as.data.frame(
    cancerprof::incidence_cancer(
      "WA", "hsa",
      cancer,
      race,
      sex,
      age,
      stage,
      year
    )
  )
  
  return(out)
}
```

```{r}
# Set parameters
race = "All Races (includes Hispanic)"
sex = "both sexes"
age = "all ages"
stage = "all stages"
year = "latest 5 year average"

cancer_types = c(
  "all cancer sites",
  "bladder",
  "brain & ons",
  "colon & rectum",
  "esophagus",
  "kidney & renal pelvis",
  "leukemia",
  "liver & bile duct",
  "lung & bronchus",
  "melanoma of the skin",
  "non-hodgkin lymphoma",
  "oral cavity & pharynx",
  "pancreas",
  "stomach",
  "thyroid"
)
```

Test code

```{r}
cancerprof::incidence_cancer(
  area = "wa",
  areatype = "county",
  cancer = "all cancer sites",
  race = "black (non-hispanic)",
  sex = "both sexes",
  age = "ages 65+",
  stage = "all stages",
  year = "latest 5 year average"
)
```


```{r}
test_data = cancerprof::incidence_cancer(
      "USA", "state",
      "bladder",
      "All Races (includes Hispanic)",
      "both sexes",
      "all ages",
      "all stages",
      "latest 5 year average"
    )

test_data
```

```{r}
for (cancer in cancer_types) {
  table_name = get_incidence_db_name(cancer, race, sex, age, stage, year)
  if (table_name %in% dbListTables(con) == FALSE) {
    dbWriteTable(
      con,
      table_name,
      get_incidence_df(cancer, race, sex, age, stage, year)
    )
  }
}
```

```{r}
dbListTables(con)
```

# Merge DB data

```{r}
# Set parameters
race = "All Races (includes Hispanic)"
sex = "both sexes"
age = "all ages"
stage = "all stages"
year = "latest 5 year average"

cancer_types = c(
  "all cancer sites",
  "bladder",
  "brain & ons",
  "colon & rectum",
  "esophagus",
  "kidney & renal pelvis",
  "leukemia",
  "liver & bile duct",
  "lung & bronchus",
  "melanoma of the skin",
  "non-hodgkin lymphoma",
  "oral cavity & pharynx",
  "pancreas",
  "stomach",
  "thyroid"
)
```

```{r}
# Get relevant tables from the db related to the parameters of interest
relevant_tables = c()
for (cancer in cancer_types) {
  table_name = get_incidence_db_name(cancer, race, sex, age, stage, year)
  if (table_name %in% dbListTables(con)) {
    relevant_tables = c(relevant_tables, table_name)
  }
}
```

```{r}
# Merge relevant tables to get final output data for cartographic visualization
incidence_by_cancer_type = data.frame()
for (tableName in relevant_tables) {
  cancerName = strsplit(tableName, "_")[[1]][3]
  if (nrow(incidence_by_cancer_type) == 0) {
    incidence_by_cancer_type = dbReadTable(con, tableName) %>%
      rename(!!cancerName := Age_Adjusted_Incidence_Rate) %>%
      select(Health_Service_Area, !!cancerName)
  } else {
    incidence_by_cancer_type <- incidence_by_cancer_type %>%
      left_join(
        dbReadTable(con, tableName) %>%
        rename(!!cancerName := Age_Adjusted_Incidence_Rate) %>%
        select(Health_Service_Area, !!cancerName), 
        by = "Health_Service_Area"
      )
  }
}

incidence_by_cancer_type
```

```{r}
dbDisconnect(con)
```

# Visualize data (TBD)

```{r}
hrr_shapefile <- st_read("./HSA_Bdry__AK_HI_unmodified/hsa-shapefile/HsaBdry_AK_HI_unmodified.shp")
```

```{r}
head(hrr_shapefile)
```

https://doh.wa.gov/data-statistical-reports/data-systems/geographic-information-system/downloadable-data-sets

https://www.cdc.gov/nchs/data/misc/atlasmet.pdf

https://geo.wa.gov/datasets/12712f465fc44fb58328c6e0255ca27e/explore

https://statecancerprofiles.cancer.gov/map/map.withimage.php?53&hsa&001&001&00&0&01&0&1&5&0#results

Going with: https://data.dartmouthatlas.org/supplemental/

# Searching for parameters that can produce cancer incidence data

Caveats: - For cancer types: 'breast (female)', 'breast (female in situ)', 'cervix', and 'uterus (corpus & uterus, nos)', we must specify that the sex is 'females' only - For cancer types: 'childhood (ages \<15, all sites)', we must specify that the age is 'ages \<15' - For cancer types: 'childhood (ages \<20, all sites)', we must specify that the age is 'ages \<20' - For cancer types: 'prostate', we must specify that the sex is 'males' only

Let's focus on just cancer type for now. We can incorporate sex, race, age, cancer stage, and year later on. Let's also assume we're sticking with the same area and areatype (WA, hsa). Lastly, we ignore childhood cancer types (b/c these are age-dependent) as well as breast, cervical, uterine, and prostate cancer types (b/c these are sex-dependent)

Let's start by saving out the incidence for different cancer types, for all sexes, ages, and races, as well as for all stages and the latest 5-year average.

## Area Type

-   County never seems to produce output for "area type" (regardless of area)
-   Works if we go with HSA instead of County for the area type.
